\bf A note on the instability and degeneracy of deep learning models

### 2017-09-30 writing/paper-note/docs/note.tex
what a mess

\usepackage{tikz, subfig, amsthm, multirow, float}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning}
\usepackage{bm}

%%INCLUDED-AK
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

@@

@@

@@

We consider the behavior, and the potential impropriety, of probability
models built to incorporate a sequence of discrete observations with
length \(N\). Let \((X_1, \dots, X_N)\) denote a set of discrete random
variables with a finite sample space, \(\mathcal{X}^N\). That is,
\(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) represents a finite set
of potential outcomes for each single variable \(X_i\), and the data
sequence \((X_1,\ldots,X_N)\) takes values in the \(N\)-fold product
space \(\mathcal{X}^N\). For each \(N\), let
\(P_{\boldsymbol \theta_N}\) denote a probability model on
\(\mathcal{X}^N\), under which
\(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of
the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we
assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
sample space \(\mathcal{X}^N\). This framework produces a series
\(P_{\boldsymbol \theta_N}\) of probability models, indexed by a generic
sequence of parameters \(\boldsymbol \theta_N\), to describe data of
each length \(N \geq 1\). The size and structure of such parameters are
without restriction, and natural cases include those where
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
integer-valued function \(q(\cdot) \geq 1\). We will refer to this model
class as \emph{Finitely Supported Finite Sequence (FSFS) models}.

Section \ref{examples} provides several examples of FSFS models commonly
used in graph/network analysis and machine learning (i.e., deep learning
models). Section \ref{instability-results} establishes formal results
regarding the propriety of FSFS models with regard to stability. A FSFS
probability model sequence exhibits instability if small changes in the
components of a data outcome \((x_1,\ldots,x_N)\) can result in large
changes in probability \(P_{\theta_N}(x_1,\ldots,x_N)\). The concept of
instability, introduced in the field of statistical physics by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}), was extended to
include a notion of detection and quantification for certain exponential
family models by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}). For
similar exponential models, particularly in connection to random
graphs/networks, Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) considered
(mean-based) characterizations for so-called model degeneracy, whereby a
probability model places all mass on a small subset of the sample space
and produces undesirably low variability in model outcomes. As described
by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), model
instability and model degeneracy are related by viewing degeneracy as an
extreme or limiting form of instability. The instability results of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) were
developed for the case of discrete exponential family models. The main
results here concern a general measure of model instability, appropriate
across the whole FSFS model class. This can be used to identify when
certain maximal probabilities in FSFS models are too extreme relative to
the length \(N\) and may thereby induce a potentially undesirable
probability structure. In this case, a one component change in the data
sequence may shift probability by orders of magnitude, and FSFS models
are rigorously shown to become degenerate as the measure of instability
increases. Lastly, Section \ref{implications} emphasizes the
implications of our model propriety results and proofs of the main
results appear in Appendix \ref{appendix-instab}.

@@

For discrete random variables \(\boldsymbol X = (X_1, \dots, X_N)\) with
sample space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
exponential family model for \(\boldsymbol X\) with probability mass
function of the form \[
p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\] depending on parameter vector
\(\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}\) and natural
parameter function
\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
positive integers \(k\) and \(L\) denoting their dimensions. Above,
\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
sufficient statistics, while \[
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
\] denotes the normalizing function, and
\(\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}\)
is the parameter space.

Defining
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)\)
with \(\boldsymbol \theta_N=\boldsymbol \lambda_N\) to be a sequence of
elements of \(\Lambda \subset \mathbb{R}^k\) and noting that
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), these discrete exponential family
models are special cases of the FSFS models. Such exponential models
arise with spatial data on a lattice (Besag
\protect\hyperlink{ref-besag1974spatial}{1974}), network data (Wasserman
and Faust \protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}), and even standard
independence models for discrete data, such as with \(N\) iid Bernoulli
random variables. (Note that for random graphs or networks with, say,
\(m\) nodes, one may wish to consider \(N={m \choose 2}\) edges as
binary (presence/absence) variables \(X_i\). In this case, the length
\(N\) of data sequence may naturally increase as a function of \(m\).)
For these exponential models, the dimension of the parameter
\(\boldsymbol \theta_N\) remains constant over each \(N\), as
\(\boldsymbol \theta_N\) lies in a parameter space of fixed Euclidean
dimension \(k\). This need not be true for other types of FSFS models
considered in Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning}. Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered
instability in such exponential models (e.g., for random graphs) for
sequences of fixed parameters
\(\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k\),
\(N \geq 1\).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, with binary
variables being most common (cf. Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
architecture has two layers, hidden (\(\mathcal{H}\)) and visible
(\(\mathcal{V}\)), with conditional independence within each layer. Let
\(\boldsymbol X=(X_1,\ldots,X_N)\) denote the \(N\) random variables for
visibles with support \(\mathcal{X}^N\) and
\(\boldsymbol H=(H_1,\ldots,H_{N_H})\) denote the \(N_H\) random
variables for hiddens with support \(\mathcal{X}^{N_H}\) where
\(\mathcal{X} = \{-1,1\}\). For parameters
\(\boldsymbol \alpha \in \mathbb{R}^{N_H}\),
\(\boldsymbol \beta \in \mathbb{R}^N\), and \(\Gamma\) as a matrix with
dimension \(N_H \times N\), the RBM model for
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) then has the
joint probability mass function \[
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
\] where \[
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
\] is the normalizing function. Let
\(\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}\)
with \(q(N) = N + N_H + N*N_H\) denote the vector of parameters for the
RBM. The probability mass function for the visible variables
\(X_1, \dots, X_N\) follows from marginalizing the joint specification
\[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
\] Note that the vector of model parameters \(\boldsymbol \theta_N\), of
size \(q(N)\), grows in size as a function of sample dimension \(N\) to
accommodate the dimension of visible variables \(X_1, \dots, X_N\), and
one may further choose the number \(N_H\) of hidden variables to change
with \(N\) as well. In particular, the number \(N_H\) of hiddens may
also potentially and arbitrarily increase with \(N\). Additionally, as
\(|\mathcal{X}| = 2\) and
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), the RBM model specification for
visibles is a FSFS model. This example also indicates that models formed
by marginalizing a base FSFS model (e.g., a type of exponential family
model) is again a FSFS model class.

@@

Consider two models with ``deep architecture'' that contain multiple
hidden (or latent) layers in addition to a visible layer of data, namely
a deep Boltzmann machine (Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}) and a deep belief
network (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). Let \(M\) denote the
number of hidden layers included in the model and let
\(N_{(H,1)}, \dots, N_{(H,M)}\) denote the numbers of hidden variables
within each hidden layer. Then the random vector
\(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\)
collects both the hidden variables
\(\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}\) and visible
variables \(\boldsymbol X =(X_1,\ldots,X_N)\) in a deep probabilistic
model. Each variable outcome will again lie in
\(\mathcal{X} = \{-1,1\}\).

>>>>

\usepackage{tikz, subfig, amsthm, multirow, float, enumerate}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning}
\usepackage{bm}

%%INCLUDED-AK
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

@@

\newcommand{\REP}{\mathrm{LREP}}
\newcommand{\DN}{\Delta_N}

\newcommand{\ma}{\mathrm{max}_{\boldsymbol \theta_N}}
\newcommand{\mi}{\mathrm{min}_{\boldsymbol \theta_N}}

\newcommand{\maa}{\mathrm{max}_{-\boldsymbol \theta_N}}
\newcommand{\mii}{\mathrm{min}_{-\boldsymbol \theta_N}}

@@

\newcommand{\elt}{A_{N}(\boldsymbol \theta_N) }
\newcommand{\Gam}{B_{N}(\boldsymbol \theta_N) }

\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }

@@

We consider the behavior, and the potential impropriety, sequences of
discrete probability models built to incorporate observations of
increasing sample size \(N\). Interest is in identifying instability in
such models, which is roughly characterized by probabilities with
extreme sensitivity to small changes in data configuration. The concept
of instability was introduced in the field of statistical physics (i.e.,
point processes) by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
extended by Schweinberger Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
family of exponential models. At issue, models exhibiting instability
are typically undesirable as these tend to provide poor representations
of data or data-generation. As an example, such models can include
near-degenerate distributions that assign all essential probability mass
to only a subset of an overall sample space. The latter issue in
connection to degeneracy has been recognized as concern in that dominant
model outcomes may not resemble observed data (cf. Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
issue, model instability often has direct negative impacts for
statistical inference and computations based on likelihood functions.
Namely, volatilities in probability structure can potentially hamper the
numerical evaluations required for maximum likelihood estimation as well
as other model-based simulations via Markov Chain Monte Carlo (MCMC).
These reasons motivate our general study of instability for a broad
class of probability models, described next.

In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\)
denote a collection of discrete random variables with a finite sample
space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian
product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\)
denotes the set of potential outcomes for each single variable \(X_i\),
so that the product space \(\mathcal{X}^N\) corresponds to values for
the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let
\(P_{\boldsymbol \theta_N}\) denote a probability model on
\(\mathcal{X}^N\), under which
\(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of
the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we
assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
sample space \(\mathcal{X}^N\). This framework produces probability
models \(P_{\boldsymbol \theta_N}\), indexed by a generic sequence of
parameters \(\boldsymbol \theta_N\), to describe data
\(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For
simplicity, we will refer to this distributional class as \emph{Finite
Outcome Everywhere Supported (FOES)} models in the following. The
dimension and structure of such parameters are generic, without
restriction, though natural cases will be seen to include those where
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
integer-valued function \(q(\cdot) \geq 1\).

Section \ref{examples} provides some examples of FOES models encountered
in graph/network analysis and machine learning (i.e., deep learning
models). These are used as references for later illustrations. Section
\ref{instability-results} then establishes several formal results for
FOES models with regard to instability. Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
developed instability results framed for, or specific to, a certain
class of discrete exponential models. For similar exponential models
with random networks, Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) studied model
degeneracy, where a probability model places near complete mass on modes
and may thereby narrow the feasible model outcomes. As findings here and
from Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
model instability and degeneracy may also be related by viewing
degeneracy as an extreme, or limiting form, of instability. Our main
results establish a broad characterization of model instability,
appropriate across the whole FOES model class, that incorporates results
of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a
special case. We prescribe a general and simple condition for
identifying instability in a FOES model sequence, which quantifies
whether certain maximal probabilities in a FOES model are too extreme
relative to the sample size \(N\). When fulfilled, the probability
structure of a FOES model is shown to exhibit extreme sensitivity, with
probability assignments possessing extreme peaks and troughs across
nearly identical outcomes. As the measure of model instability
increases, probabilities from an unstable FOES model additionally
increase in volatility and provably slide into degeneracy. Section
\ref{implications} then emphasizes the implications of such model
instability, showing that such impropriety can be expected to
numerically hinder maximum likelihood estimation and MCMC-based
simulations. As one potential remedy, suggestions are given or
constraining model parameterizations and avoid the most problematic
regions of the parameter space. Proofs of the main results appear in
Appendix \ref{appendix-instab}.

@@

For random variables
\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
exponential family model for \(\boldsymbol X\) with probability mass
function given by
\begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector
\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
parameter function
\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
positive integers \(k\) and \(L\) denoting their dimensions. Above,
\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
sufficient statistics, while \[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\] denotes the normalizing function with parameter space \(\Theta_N\).
The natural parameter function \(\eta (\boldsymbol \theta)\) has a
linear form (i.e.,
\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
\(L \times k\) matrix \(\bm{A}\)) in many common model formulaitons,
though may also be nonlinear (e.g., curved exponential families). In the
linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
generally assumed in the exponential parameterization with a minor
modification to the definition of sufficient statistics
\(\boldsymbol g_N(\boldsymbol x)\).

Such discrete exponential family models are special cases of the FOES
models, as seen by defining
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\),
\(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a
parameter sequence
\(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example,
if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\)
independent and identically distributed Bernoulli random variables, each
indicating a binary \(0\)-\(1\) outcome, the resulting probabilities
have exponential form \eqref{eq:expo} given by
\begin{equation}
\label{eq:mod1}
P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
 \end{equation}
with sufficient statistic
\(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds
ratio'' parameter
\(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent
\(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\)
among \(k\) possibilities (e.g., a die roll), a multinomial distribution
is given by
\begin{equation}
\label{eq:mod11}
P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
\end{equation}
with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a
count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
\(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the
indicator function, and parameters
\(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\)
defining log-probability ratios
\(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
In addition to such standard models for discrete independent data,
exponential models of FOES type commonly arise with dependent spatial
data (Besag \protect\hyperlink{ref-besag1974spatial}{1974}) and
network/relational data (Wasserman and Faust
\protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For a random graph
or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random
edges where the \(i\)th edge is associated with a pair of nodes
\(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\)
indicating presence/absence of an edge among the node pair \(s_i\),
\(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence
\(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node
number \(n\) and corresponding exponential models often incorporate
graph topographical features derived from \(\boldsymbol X\). As an
example, consider a graph model of exponential/FOES form prescribed by
\begin{equation}
\label{eq:mod2}
P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
\end{equation}
\[
  g_{1,N}(\boldsymbol x) \equiv \sum_{i=1}^N  x_i, \qquad\quad g_{2,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j \leq N,\atop s_i \cap s_j \neq \emptyset}\!\!\!   x_i x_j, \qquad
  g_{3,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j<\ell \leq N, \atop s_i \cap s_j \neq \emptyset,s_i \cap s_\ell \neq \emptyset, \\ s_j \cap s_\ell \neq \emptyset } \!\!\!\!\!\!\!\!  \!\!\!\!\!\!\!\!   x_i x_j x_\ell,
\] involving the numbers of edges, 2-stars and triangles among an
outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\),
\(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively,
along with \(k=3\) real parameters
\(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
For this network model \eqref{eq:mod2} in particular, as well as for more
general models of form \eqref{eq:expo}, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered
instability in such exponential models with sequences of fixed
parameters
\(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\),
\(N \geq 1\), of fixed dimension \(k\).

For model sequences
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\)
of the exponential type \eqref{eq:expo}, such as those in
\eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the
parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\)
necessarily remains the same for all sample sizes \(N \geq 1\) as the
form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo}
and the number of sufficient statistics
\(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of
fixed Euclidean dimension \(k\). However, this aspect need not be true
for other types of FOES models considered in Sections
\ref{restricted-boltzmann-machines} - \ref{deep-learning}, where instead
the numbers of parameters and sufficient statistics commonly increase
with the sample size \(N\).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, with binary
variables being most common (cf. Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
architecture has two layers, hidden (\(\mathcal{H}\)) and visible
(\(\mathcal{V}\)), with conditional independence within each layer. Let
\(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables
for visibles with support \(\mathcal{X}^N\) and
\(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the
\(N_\mathcal{H}\) random variables for hiddens with support
\(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For
parameters
\(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\),
\(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and
\(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension
\(N_\mathcal{H} \times N\), the RBM model for
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint
probability mass function
\begin{equation}
\label{eq:RBM1}
\tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
\end{equation}
with normalizing function \[
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
\] Let
\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\),
with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote the
parameter vector for the RBM, as indexed by the number \(N\) of visible
random variables (which may differ from the actual lengths of these
parameter vectors). The probability mass function for the visible
variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from
marginalizing the joint specification to yield
\begin{equation}
\label{eq:RBM2}
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
\end{equation}
Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a
linear exponential one in sufficient statistics
\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from
\eqref{eq:RBM1}, but the form differs from the previous exponential models
in \eqref{eq:expo} in that the lengths of parameters
\(\boldsymbol \theta_N\) and statistics
\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to
incorporate more visible variables. That is, in contrast to
\eqref{eq:expo}, the natural parameter function involved in the RBM model
\eqref{eq:RBM1}, as the identity mapping of the parameters
\(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in
dimension \(q(N)\to \infty\) to accommodate visible variables
\(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
Additionally, one may further arbitrarily choose the number
\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RMB
model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the
\(N\) visible variables \(\boldsymbol X\), and the number
\(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
Because \(|\mathcal{X}| = 2\) and
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2}
for visibles \(\boldsymbol\) corresponds to a FOES model, while the
joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\)
is also a FOES model. As this example also indicates, any model formed
by marginalizing a base FOES model class, such as the RBM joint
specification \eqref{eq:RBM1}, is again a FOES model.

@@

Consider two models with ``deep architecture'' that contain multiple
hidden (or latent) layers in addition to a visible layer of data, namely
a deep Boltzmann machine (Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}) and a deep belief
network Hinton, Osindero, and Teh
(\protect\hyperlink{ref-hinton2006fast}{2006}){]}. Let \(M\) denote the
number of hidden layers included in the model and let
\(N_{(H,1)}, \dots, N_{(H,M)}\) denote the numbers of hidden variables
within each hidden layer. Then the random vector
\(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\)
collects both the hidden variables
\(\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}\) and visible
variables \(\boldsymbol X =(X_1,\ldots,X_N)\) in a deep probabilistic
model. Each variable outcome will again lie in
\(\mathcal{X} = \{-1,1\}\).

\textbf{Deep Boltzmann machine (DBM).} The DBM class of models maintains
conditional independence within all layers in the model by stacking RBM
models and only allowing conditional dependence between neighboring
layers. The joint probability mass function for a DBM is \[
\tilde{P}_{\boldsymbol \theta_N} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
\] for
\(\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}\)
where \[
\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
\] is the normalizing function for
\(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\),
consisting of model parameters \(\boldsymbol \beta \in \mathbb{R}^N\),
\(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\),
\(i = 1, \dots, M\), along with a matrix \(\Gamma^{(0)}\) of dimension
\(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\). The combined
parameter vector \(\boldsymbol \theta_N\) has total length
\(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).
The probability mass function for the visible random variables
\(X_1, \dots, X_N\) follows from this joint specification as \[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
\] Again like the RBM case, the DBM model specification is an example of
a FOES model.
